---
title: "Adult Children Caregivers Analysis"
author: "Anna Bokun"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
    df_print: paged
  word_document:
    toc: true
    reference_docx: word-styles-reference.docx
  pdf_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

# Load required libraries
library(dplyr)
library(knitr)
library(tidyr)
library(kableExtra)
library(ggplot2)
library(scales)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                     fig.width = 10, fig.height = 7)

# Read the data
# Adjust the path to your data file
ACC <- read.csv("/Users/annabokun/Google Drive/My Drive/UT-Austin Postdoc/Projects/Caregivers - Housing Health Paradox/Replication Files/Data/ACC.csv")
```

## Introduction

This document analyzes the characteristics of Mexican-American adult child caregivers from the Hispanic Established Populations for Epidemiologic Study of the Elderly (HEPESE) Wave 7 dataset (2010-2011). The analysis focuses specifically on adult children who provide care to their aging parents, examining different groups based on co-residence and household headship patterns.

## Methodology

We examine three distinct groups of adult children caregivers:

1. **Co-residing adult children caregivers**: Adult children who live in the same household as their parent care recipients
2. **Adult children who moved in with parents (general)**: Adult children who moved in with their parents without specifying caregiving as the primary reason
3. **Adult children who moved in for caregiving**: Adult children who moved in specifically to provide care to their parents

For each group, we analyze demographic characteristics, socioeconomic status, health status, caregiving intensity, and financial strain.

```{r data-prep}
# Filter for adult children caregivers
adult_children <- ACC %>%
  filter(RELSUBJ7I %in% c(3, 4))  # 3 = Child, 4 = Child-in-law

# TABLE 1: All co-residing adult children caregivers
coresiding <- adult_children %>%
  filter(cg_coreside == 1)

# TABLE 2: Adult children caregivers who moved in with parents (general)
moved_in_general <- adult_children %>%
  filter(move_in_type == 5)

# TABLE 3: Adult children caregivers who moved in for caregiving
moved_in_care <- adult_children %>%
  filter(move_in_type == 2)

# Create a function to calculate summary statistics
create_summary_table <- function(data_group, group_name) {
  
  # Sample size
  n_total <- nrow(data_group)
  if(n_total == 0) {
    return(data.frame(
      characteristic = "No data available",
      level = "",
      count = 0,
      percentage = NA
    ))
  }
  
  # Gender
  gender <- data_group %>%
    group_by(gender = ifelse(cg_female == 1, "Female", "Male")) %>%
    summarise(count = n()) %>%
    mutate(percentage = round(count/sum(count)*100, 1),
           characteristic = "Gender",
           level = gender) %>%
    select(characteristic, level, count, percentage)
  
  # Age
  age <- data_group %>%
    group_by(age = cg_age_cat) %>%
    summarise(count = n()) %>%
    filter(!is.na(age)) %>%
    mutate(percentage = round(count/sum(count)*100, 1),
           characteristic = "Age group",
           level = age) %>%
    select(characteristic, level, count, percentage) %>%
    arrange(match(level, c("18-44", "45-54", "55-64", "65-74")))
  
  # Education
  education <- data_group %>%
    group_by(education = cg_educ_cat) %>%
    summarise(count = n()) %>%
    filter(!is.na(education)) %>%
    mutate(percentage = round(count/sum(count)*100, 1),
           characteristic = "Education",
           level = education) %>%
    select(characteristic, level, count, percentage) %>%
    arrange(match(level, c("0-6 years", "7-11 years", "High school", "Bachelor's+")))
  
  # Income
  income <- data_group %>%
    group_by(income = cg_inc_cat) %>%
    summarise(count = n()) %>%
    filter(!is.na(income)) %>%
    mutate(percentage = round(count/sum(count)*100, 1),
           characteristic = "Income",
           level = income) %>%
    select(characteristic, level, count, percentage) %>%
    arrange(match(level, c("<$10k", "$10-19k", "$20-49k", "$50k+")))
  
  
  # Medicaid status
  medicaid <- data_group %>%
    group_by(medicaid = ifelse(cg_medicaid_dum == 1, "On Medicaid", "Not on Medicaid")) %>%
    summarise(count = n()) %>%
    filter(!is.na(medicaid)) %>%
    mutate(percentage = round(count/sum(count)*100, 1),
           characteristic = "Medicaid status",
           level = medicaid) %>%
    select(characteristic, level, count, percentage)

  # Marital status
  marital <- data_group %>%
    group_by(marital = ifelse(cg_married == 1, "Married", "Not married")) %>%
    summarise(count = n()) %>%
    filter(!is.na(marital)) %>%
    mutate(percentage = round(count/sum(count)*100, 1),
           characteristic = "Marital status",
           level = marital) %>%
    select(characteristic, level, count, percentage)
  
  # Health status
  health <- data_group %>%
    group_by(health = cg_health) %>%
    summarise(count = n()) %>%
    filter(!is.na(health)) %>%
    mutate(percentage = round(count/sum(count)*100, 1),
           characteristic = "Health status",
           level = health) %>%
    select(characteristic, level, count, percentage)
  
  # Financial strain
  financial <- data_group %>%
    group_by(strain = ifelse(cg_finc_strain == 1, "Has financial strain", "No financial strain")) %>%
    summarise(count = n()) %>%
    filter(!is.na(strain)) %>%
    mutate(percentage = round(count/sum(count)*100, 1),
           characteristic = "Financial strain",
           level = strain) %>%
    select(characteristic, level, count, percentage)
  
  # Caregiving intensity
  intensity <- data_group %>%
    group_by(intensity = cg_intensity) %>%
    summarise(count = n()) %>%
    filter(!is.na(intensity)) %>%
    mutate(percentage = round(count/sum(count)*100, 1),
           characteristic = "Caregiving intensity",
           level = intensity) %>%
    select(characteristic, level, count, percentage) %>%
    arrange(match(level, c("Low", "Moderate", "High")))
  
  # Household headship
  hoh <- data_group %>%
    group_by(hoh = ifelse(cg_HOH == 1, "Head of household", "Not head of household")) %>%
    summarise(count = n()) %>%
    filter(!is.na(hoh)) %>%
    mutate(percentage = round(count/sum(count)*100, 1),
           characteristic = "Household headship",
           level = hoh) %>%
    select(characteristic, level, count, percentage)
  
  # Mean ADL and IADL hours
  adl_hours <- mean(data_group$cg_hours_adl, na.rm = TRUE)
  iadl_hours <- mean(data_group$cg_hours_iadl, na.rm = TRUE)
  
  caregiving_hours <- data.frame(
    characteristic = c("Caregiving hours", "Caregiving hours"),
    level = c("Mean ADL hours/day", "Mean IADL hours/day"),
    count = c(round(adl_hours, 2), round(iadl_hours, 2)),
    percentage = NA
  )
  
  # Check if the parent has dementia
  if("cr_dementia" %in% names(data_group)) {
    dementia <- data_group %>%
      filter(!is.na(cr_dementia_binary)) %>%
      group_by(dementia = ifelse(cr_dementia_binary == 1, "Parent has dementia", "Parent does not have dementia")) %>%
      summarise(count = n()) %>%
      mutate(percentage = round(count/sum(count)*100, 1),
             characteristic = "Parent cognitive status",
             level = dementia) %>%
      select(characteristic, level, count, percentage)
  } else {
    dementia <- data.frame(
      characteristic = "Parent cognitive status",
      level = "Data not available",
      count = NA,
      percentage = NA
    )
  }
  
  # Combine all summaries
  summary_table <- bind_rows(
    gender, age, education, income, medicaid, marital, health, financial, intensity, hoh, caregiving_hours, dementia
  )
  
  # Add group name and sample size
  attr(summary_table, "group_name") <- group_name
  attr(summary_table, "sample_size") <- n_total
  
  return(summary_table)
}

# Create the three tables
table1 <- create_summary_table(coresiding, "All Co-residing Adult Children Caregivers")
table2 <- create_summary_table(moved_in_general, "Adult Children Who Moved In with Parents (General)")
table3 <- create_summary_table(moved_in_care, "Adult Children Who Moved In for Caregiving")
```

## Results

### Table 1: Characteristics of All Co-residing Adult Children Caregivers

```{r table1}
# Format and display table with kable and kableExtra
kable(table1, 
      col.names = c("Characteristic", "Category", "n", "%"),
      caption = paste("All Co-residing Adult Children Caregivers (N =", attr(table1, "sample_size"), ")")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = TRUE, 
                position = "center",
                font_size = 12) %>%
  column_spec(1, bold = TRUE) %>%
  collapse_rows(columns = 1, valign = "top") %>%
  row_spec(0, bold = TRUE, background = "#f8f8f8") %>%
  footnote(general = "Data source: HEPESE Wave 7 (2010-2011)")
```

### Table 2: Characteristics of Adult Children Who Moved In with Parents (General)

```{r table2}
kable(table2, 
      col.names = c("Characteristic", "Category", "n", "%"),
      caption = paste("Adult Children Who Moved In with Parents - General (N =", attr(table2, "sample_size"), ")")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = TRUE, 
                position = "center",
                font_size = 12) %>%
  column_spec(1, bold = TRUE) %>%
  collapse_rows(columns = 1, valign = "top") %>%
  row_spec(0, bold = TRUE, background = "#f8f8f8") %>%
  footnote(general = "Data source: HEPESE Wave 7 (2010-2011)")
```

### Table 3: Characteristics of Adult Children Who Moved In for Caregiving

```{r table3}
kable(table3, 
      col.names = c("Characteristic", "Category", "n", "%"),
      caption = paste("Adult Children Who Moved In for Caregiving (N =", attr(table3, "sample_size"), ")")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = TRUE, 
                position = "center",
                font_size = 12) %>%
  column_spec(1, bold = TRUE) %>%
  collapse_rows(columns = 1, valign = "top") %>%
  row_spec(0, bold = TRUE, background = "#f8f8f8") %>%
  footnote(general = "Data source: HEPESE Wave 7 (2010-2011)")
```

## Comparing Financial Strain Across Groups

```{r financial-strain, fig.width=10, fig.height=6}
# Extract financial strain data for visualization
extract_financial_strain <- function(table_data) {
  group_name <- attr(table_data, "group_name")
  
  strain_data <- table_data %>%
    filter(characteristic == "Financial strain") %>%
    mutate(group = group_name)
  
  return(strain_data)
}

# Combine financial strain data from all tables
financial_strain_data <- bind_rows(
  extract_financial_strain(table1),
  extract_financial_strain(table2),
  extract_financial_strain(table3)
)

# Plot financial strain comparison
ggplot(financial_strain_data %>% filter(level == "Has financial strain"), 
       aes(x = reorder(group, percentage), y = percentage, fill = group)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(aes(label = paste0(percentage, "%")), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5, 
            size = 4) +
  ylim(0,100) +
  labs(title = "Financial Strain Among Different Groups of Adult Children Caregivers",
       subtitle = "Percentage reporting financial strain",
       x = "",
       y = "Percentage") +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.title.y = element_text(size = 12),
    aspect.ratio = 0.6, 
    panel.grid.major.x = element_blank()
  )
```



## Comparing Caregiving Hours Across Groups

```{r caregiving-hours, fig.width=10, fig.height=6}
# Extract caregiving hours data for visualization
extract_caregiving_hours <- function(table_data) {
  group_name <- attr(table_data, "group_name")
  
  hours_data <- table_data %>%
    filter(characteristic == "Caregiving hours") %>%
    mutate(group = group_name,
           hours = count) %>%
    select(group, level, hours)
  
  return(hours_data)
}

# Combine caregiving hours data from all tables
caregiving_hours_data <- bind_rows(
  extract_caregiving_hours(table1),
  extract_caregiving_hours(table2),
  extract_caregiving_hours(table3)
)

# Convert to long format for plotting
caregiving_hours_long <- caregiving_hours_data %>%
  pivot_wider(names_from = level, values_from = hours) %>%
  rename(ADL = `Mean ADL hours/day`, IADL = `Mean IADL hours/day`) %>%
  pivot_longer(cols = c(ADL, IADL), names_to = "hours_type", values_to = "hours")

# Plot caregiving hours comparison
ggplot(caregiving_hours_long, aes(x = group, y = hours, fill = hours_type)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  geom_text(aes(label = round(hours, 1)), 
            position = position_dodge(width = 0.7), 
            vjust = -0.5, 
            size = 4) +
  ylim(0,10) +
  labs(title = "Caregiving Hours Among Different Groups of Adult Children Caregivers",
       subtitle = "Mean hours per day spent on ADL and IADL assistance",
       x = "",
       y = "Hours per day",
       fill = "Care Type") +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.title.y = element_text(size = 12),
    aspect.ratio = 0.6, 
    panel.grid.major.x = element_blank()
  )
```

## Discussion

This analysis reveals several important patterns among adult children caregivers:

1. **Financial strain:** Adult children who moved in specifically for caregiving reasons experience higher rates of financial strain compared to caregivers who co-reside without discerning who moved in with whom (i.e., Table 1) (22.2% vs 15.7%). 

2. **Caregiving intensity:** The data shows variation in caregiving intensity across different groups. Those who moved in specifically for caregiving spent more hours on caregiving tasks, relative to caregivers who co-reside without discerning who moved in with whom (8.04-8.05 ADL/IADL hours vs 6.21-6.29 ADL/IADL hours).

3. **Demographic differences:** There are notable demographic differences between adult children who moved in specifically for caregiving versus those who co-reside for other reasons, particularly in terms of gender distribution, marital status, and economic resources. For example, 11.1% of adult children who moved in for caregiving reasons report participating in Medicaid, vs 21.1% who moved in more generally, and 16.8% for those who co-reside without discerning who moved in with whom.  

4. **Household headship:** Among adult children who moved in specifically for caregiving reasons, 85.2% are not the HOH vs 42.9% for adult children who moved in for general reasons. The pattern of household headship varies across groups, which may have implications for financial decision-making, resource allocation, and caregiver autonomy within the household.  


## Limitations

Several limitations should be noted when interpreting these results:

1. The data is cross-sectional, limiting our ability to conduct analyses over time and to establish causal relationships between co-residence patterns and financial strain.

2. The sample sizes for some subgroups are prohibitively small (n < 30), which affects the reliability of estimates.

3. The HEPESE data is specific to Mexican-Americans in southwestern states and may not generalize to other Latino populations or regions.

4. Self-reported measures of financial strain and living arrangements may be subject to recall bias or social desirability effects.
