---
title: "Adult Children Caregivers Analysis"
author: "HEPESE Data Analysis"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 2
  word_document:
    toc: true
    reference_docx: word-styles-reference.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                     fig.width = 10, fig.height = 7) # Default figure dimensions

# Load required libraries
library(dplyr)
library(knitr)
library(tidyr)
library(kableExtra)
library(ggplot2)
library(scales)
library(purrr) # Add purrr for map functions

# Read the data
# Adjust the path to your data file
ACC <- read.csv("/Users/annabokun/Google Drive/My Drive/UT-Austin Postdoc/Projects/Caregivers - Housing Health Paradox/Replication Files/Data/ACC.csv")
```

## Introduction

This document analyzes the characteristics of Mexican-American adult child caregivers from the Hispanic Established Populations for Epidemiologic Study of the Elderly (HEPESE) Wave 7 dataset (2010-2011). The analysis focuses specifically on adult children who provide care to their aging parents, examining different groups based on co-residence and household headship patterns.

## Methodology

We examine three distinct groups of adult children caregivers:

1. **Co-residing adult children caregivers**: Adult children who live in the same household as their parent care recipients
2. **Adult children who moved in with parents (general)**: Adult children who moved in with their parents without specifying caregiving as the primary reason
3. **Adult children who moved in for caregiving**: Adult children who moved in specifically to provide care to their parents

For each group, we analyze demographic characteristics, socioeconomic status, health status, caregiving intensity, and financial strain.

```{r data-prep}
# Filter for adult children caregivers
adult_children <- ACC %>%
  filter(RELSUBJ7I %in% c(3, 4))  # 3 = Child, 4 = Child-in-law

# TABLE 1: All co-residing adult children caregivers
coresiding <- adult_children %>%
  filter(coreside == 1)

# TABLE 2: Adult children caregivers who moved in with parents (general)
# Category 5: General move-in by adult child
moved_in_general <- adult_children %>%
  filter(move_in_type == 5)

# TABLE 3: Adult children caregivers who moved in for caregiving
# Category 2: Adult child likely moved in for caregiving reasons
moved_in_care <- adult_children %>%
  filter(move_in_type == 2)

# Create a function to calculate summary statistics
create_summary_table <- function(data_group, group_name) {
  
  # Sample size
  n_total <- nrow(data_group)
  if(n_total == 0) {
    return(data.frame(
      characteristic = "No data available",
      level = "",
      count = 0,
      percentage = NA
    ))
  }
  
  # Gender
  gender <- data_group %>%
    group_by(gender = ifelse(cg_female == 1, "Female", "Male")) %>%
    summarise(count = n()) %>%
    mutate(percentage = round(count/sum(count)*100, 1),
           characteristic = "Gender",
           level = gender) %>%
    select(characteristic, level, count, percentage)
  
  # Age
  age <- data_group %>%
    group_by(age = cg_age_cat) %>%
    summarise(count = n()) %>%
    filter(!is.na(age)) %>%
    mutate(percentage = round(count/sum(count)*100, 1),
           characteristic = "Age group",
           level = age) %>%
    select(characteristic, level, count, percentage) %>%
    arrange(match(level, c("18-44", "45-54", "55-64", "65-74")))
  
  # Education
  education <- data_group %>%
    group_by(education = cg_educ_cat) %>%
    summarise(count = n()) %>%
    filter(!is.na(education)) %>%
    mutate(percentage = round(count/sum(count)*100, 1),
           characteristic = "Education",
           level = education) %>%
    select(characteristic, level, count, percentage) %>%
    arrange(match(level, c("0-6 years", "7-11 years", "High school", "Bachelor's+")))
  
  # Income
  income <- data_group %>%
    group_by(income = cg_inc_cat) %>%
    summarise(count = n()) %>%
    filter(!is.na(income)) %>%
    mutate(percentage = round(count/sum(count)*100, 1),
           characteristic = "Income",
           level = income) %>%
    select(characteristic, level, count, percentage) %>%
    arrange(match(level, c("<$10k", "$10-19k", "$20-49k", "$50k+")))
  
  # Marital status
  marital <- data_group %>%
    group_by(marital = ifelse(cg_married == 1, "Married", "Not married")) %>%
    summarise(count = n()) %>%
    filter(!is.na(marital)) %>%
    mutate(percentage = round(count/sum(count)*100, 1),
           characteristic = "Marital status",
           level = marital) %>%
    select(characteristic, level, count, percentage)
  
  # Health status
  health <- data_group %>%
    group_by(health = cg_health) %>%
    summarise(count = n()) %>%
    filter(!is.na(health)) %>%
    mutate(percentage = round(count/sum(count)*100, 1),
           characteristic = "Health status",
           level = health) %>%
    select(characteristic, level, count, percentage)
  
  # Financial strain
  financial <- data_group %>%
    group_by(strain = ifelse(cg_finc_strain == 1, "Has financial strain", "No financial strain")) %>%
    summarise(count = n()) %>%
    filter(!is.na(strain)) %>%
    mutate(percentage = round(count/sum(count)*100, 1),
           characteristic = "Financial strain",
           level = strain) %>%
    select(characteristic, level, count, percentage)
  
  # Caregiving intensity
  intensity <- data_group %>%
    group_by(intensity = cg_intensity) %>%
    summarise(count = n()) %>%
    filter(!is.na(intensity)) %>%
    mutate(percentage = round(count/sum(count)*100, 1),
           characteristic = "Caregiving intensity",
           level = intensity) %>%
    select(characteristic, level, count, percentage) %>%
    arrange(match(level, c("Low", "Moderate", "High")))
  
  # Household headship
  hoh <- data_group %>%
    group_by(hoh = ifelse(cg_HOH == 1, "Head of household", "Not head of household")) %>%
    summarise(count = n()) %>%
    filter(!is.na(hoh)) %>%
    mutate(percentage = round(count/sum(count)*100, 1),
           characteristic = "Household headship",
           level = hoh) %>%
    select(characteristic, level, count, percentage)
  
  # Medicaid status
  medicaid <- data_group %>%
    group_by(medicaid = ifelse(cg_medicaid_dum == 1, "On Medicaid", "Not on Medicaid")) %>%
    summarise(count = n()) %>%
    filter(!is.na(medicaid)) %>%
    mutate(percentage = round(count/sum(count)*100, 1),
           characteristic = "Medicaid status",
           level = medicaid) %>%
    select(characteristic, level, count, percentage)
  
  # Mean ADL and IADL hours
  adl_hours <- mean(data_group$cg_hours_adl, na.rm = TRUE)
  iadl_hours <- mean(data_group$cg_hours_iadl, na.rm = TRUE)
  
  caregiving_hours <- data.frame(
    characteristic = c("Caregiving hours", "Caregiving hours"),
    level = c("Mean ADL hours/day", "Mean IADL hours/day"),
    count = c(round(adl_hours, 2), round(iadl_hours, 2)),
    percentage = NA
  )
  
  # Check if the parent has dementia
  if("cr_dementia" %in% names(data_group)) {
    dementia <- data_group %>%
      filter(!is.na(cr_dementia)) %>%
      group_by(dementia = ifelse(cr_dementia == 1, "Parent has dementia", "Parent does not have dementia")) %>%
      summarise(count = n()) %>%
      mutate(percentage = round(count/sum(count)*100, 1),
             characteristic = "Parent cognitive status",
             level = dementia) %>%
      select(characteristic, level, count, percentage)
  } else {
    dementia <- data.frame(
      characteristic = "Parent cognitive status",
      level = "Data not available",
      count = NA,
      percentage = NA
    )
  }
  
  # Combine all summaries
  summary_table <- bind_rows(
    gender, age, education, income, marital, health, medicaid, financial, intensity, hoh, caregiving_hours, dementia
  )
  
  # Add group name and sample size
  attr(summary_table, "group_name") <- group_name
  attr(summary_table, "sample_size") <- n_total
  
  return(summary_table)
}

# Create the three tables
table1 <- create_summary_table(coresiding, "All Co-residing Adult Children Caregivers")
table2 <- create_summary_table(moved_in_general, "Adult Children Who Moved In with Parents (General)")
table3 <- create_summary_table(moved_in_care, "Adult Children Who Moved In for Caregiving")
```

## Results

### Table 1: Characteristics of All Co-residing Adult Children Caregivers

```{r table1}
# Format and display table with kable and kableExtra
kable(table1, 
      col.names = c("Characteristic", "Category", "n", "%"),
      caption = paste("All Co-residing Adult Children Caregivers (N =", attr(table1, "sample_size"), ")")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = TRUE, 
                position = "center",
                font_size = 12) %>%
  column_spec(1, bold = TRUE) %>%
  collapse_rows(columns = 1, valign = "top") %>%
  row_spec(0, bold = TRUE, background = "#f8f8f8") %>%
  footnote(general = "Data source: HEPESE Wave 7 (2010-2011)")
```

### Table 2: Characteristics of Adult Children Who Moved In with Parents (General)

```{r table2}
kable(table2, 
      col.names = c("Characteristic", "Category", "n", "%"),
      caption = paste("Adult Children Who Moved In with Parents - General (N =", attr(table2, "sample_size"), ")")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = TRUE, 
                position = "center",
                font_size = 12) %>%
  column_spec(1, bold = TRUE) %>%
  collapse_rows(columns = 1, valign = "top") %>%
  row_spec(0, bold = TRUE, background = "#f8f8f8") %>%
  footnote(general = "Data source: HEPESE Wave 7 (2010-2011). Category 5: General move-in by adult child.")
```

### Table 3: Characteristics of Adult Children Who Moved In for Caregiving

```{r table3}
kable(table3, 
      col.names = c("Characteristic", "Category", "n", "%"),
      caption = paste("Adult Children Who Moved In for Caregiving (N =", attr(table3, "sample_size"), ")")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = TRUE, 
                position = "center",
                font_size = 12) %>%
  column_spec(1, bold = TRUE) %>%
  collapse_rows(columns = 1, valign = "top") %>%
  row_spec(0, bold = TRUE, background = "#f8f8f8") %>%
  footnote(general = "Data source: HEPESE Wave 7 (2010-2011). Category 2: Adult child likely moved in for caregiving reasons.")
```

## Comparing Financial Strain by Household Headship

```{r financial-strain-by-hoh, fig.width=10, fig.height=7}
# Extract and prepare financial strain data stratified by HOH status
prepare_financial_strain_data <- function(stratified_table) {
  group_name <- attr(stratified_table, "group_name")
  
  # Extract financial strain data
  fin_strain_data <- stratified_table %>%
    filter(characteristic == "Financial strain" & level == "Has financial strain") %>%
    select(hoh_status, percentage, count) %>%
    mutate(group = group_name)
  
  return(fin_strain_data)
}

# Combine financial strain data from all tables
financial_strain_by_hoh <- bind_rows(
  prepare_financial_strain_data(table1_stratified),
  prepare_financial_strain_data(table2_stratified),
  prepare_financial_strain_data(table3_stratified)
)

# Create more descriptive labels
financial_strain_by_hoh <- financial_strain_by_hoh %>%
  mutate(
    group_label = case_when(
      group == "All Co-residing Adult Children Caregivers" ~ "All Co-residing",
      group == "Adult Children Who Moved In with Parents (General)" ~ "Moved In (General)",
      group == "Adult Children Who Moved In for Caregiving" ~ "Moved In for Caregiving",
      TRUE ~ group
    ),
    combined_label = paste(group_label, hoh_status)
  )

# Plot financial strain comparison by HOH status
ggplot(financial_strain_by_hoh, 
       aes(x = reorder(combined_label, percentage), y = percentage, fill = hoh_status)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(label = paste0(percentage, "%")), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5, 
            size = 4) +
  scale_y_continuous(limits = c(0, max(financial_strain_by_hoh$percentage, na.rm = TRUE) * 1.2),
                    labels = function(x) paste0(x, "%")) +
  labs(title = "Financial Strain by Household Headship Status",
       subtitle = "Percentage reporting financial strain across caregiver groups",
       x = "",
       y = "Percentage with Financial Strain",
       fill = "Household Headship") +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.title.y = element_text(size = 12),
    panel.grid.major.x = element_blank(),
    aspect.ratio = 0.5
  )
```

## Comparing Caregiving Hours by Household Headship

```{r caregiving-hours-by-hoh, fig.width=10, fig.height=7}
# Extract and prepare caregiving hours data stratified by HOH status
prepare_caregiving_hours_data <- function(stratified_table) {
  group_name <- attr(stratified_table, "group_name")
  
  # Extract caregiving hours data
  hours_data <- stratified_table %>%
    filter(characteristic == "Caregiving hours") %>%
    mutate(
      hours_type = level,
      hours = count,
      group = group_name
    ) %>%
    select(hoh_status, hours_type, hours, group)
  
  return(hours_data)
}

# Combine caregiving hours data from all tables
caregiving_hours_by_hoh <- bind_rows(
  prepare_caregiving_hours_data(table1_stratified),
  prepare_caregiving_hours_data(table2_stratified),
  prepare_caregiving_hours_data(table3_stratified)
)

# Create more descriptive labels
caregiving_hours_by_hoh <- caregiving_hours_by_hoh %>%
  mutate(
    group_label = case_when(
      group == "All Co-residing Adult Children Caregivers" ~ "All Co-residing",
      group == "Adult Children Who Moved In with Parents (General)" ~ "Moved In (General)",
      group == "Adult Children Who Moved In for Caregiving" ~ "Moved In for Caregiving",
      TRUE ~ group
    ),
    hours_type = gsub("Mean ", "", hours_type),
    combined_label = paste(group_label, hoh_status)
  )

# Plot caregiving hours comparison by HOH status
ggplot(caregiving_hours_by_hoh, 
       aes(x = combined_label, y = hours, fill = hours_type)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  geom_text(aes(label = round(hours, 1)), 
            position = position_dodge(width = 0.7), 
            vjust = -0.5, 
            size = 3.5) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Caregiving Hours by Household Headship Status",
       subtitle = "Mean hours per day spent on ADL and IADL assistance",
       x = "",
       y = "Hours per day",
       fill = "Care Type") +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.title.y = element_text(size = 12),
    panel.grid.major.x = element_blank(),
    aspect.ratio = 0.5
  )
``` financial strain"), 
       aes(x = reorder(group, percentage), y = percentage, fill = group)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(aes(label = paste0(percentage, "%")), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5, 
            size = 4) +
  scale_y_continuous(limits = c(0, max(financial_strain_data$percentage, na.rm = TRUE) * 1.2),
                    labels = function(x) paste0(x, "%")) +
  labs(title = "Financial Strain Among Different Groups of Adult Children Caregivers",
       subtitle = "Percentage reporting financial strain",
       x = "",
       y = "Percentage") +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.title.y = element_text(size = 12),
    panel.grid.major.x = element_blank(),
    aspect.ratio = 0.6  # Adjust this value to make the plot taller
  )
```

## Comparing Medicaid Status Across Groups

```{r medicaid-plot, fig.width=10, fig.height=6}
# Create a direct data frame with Medicaid percentages
medicaid_percentages <- data.frame(
  group = c(
    attr(table1, "group_name"),
    attr(table2, "group_name"),
    attr(table3, "group_name")
  )
)

# Function to safely extract Medicaid percentage
extract_medicaid_pct <- function(table) {
  medicaid_rows <- table %>% filter(characteristic == "Medicaid status" & level == "On Medicaid")
  if(nrow(medicaid_rows) > 0) {
    return(medicaid_rows$percentage[1])
  } else {
    return(NA)
  }
}

# Function to safely extract Medicaid count
extract_medicaid_count <- function(table) {
  medicaid_rows <- table %>% filter(characteristic == "Medicaid status" & level == "On Medicaid")
  if(nrow(medicaid_rows) > 0) {
    return(medicaid_rows$count[1])
  } else {
    return(NA)
  }
}

# Add percentage and count columns
medicaid_percentages$percentage <- c(
  extract_medicaid_pct(table1),
  extract_medicaid_pct(table2),
  extract_medicaid_pct(table3)
)

medicaid_percentages$count <- c(
  extract_medicaid_count(table1),
  extract_medicaid_count(table2),
  extract_medicaid_count(table3)
)

# Create a simple bar chart with the medicaid_percentages data
ggplot(medicaid_percentages, aes(x = reorder(group, percentage), y = percentage, fill = group)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(aes(label = paste0(percentage, "%")), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5, 
            size = 4) +
  scale_y_continuous(limits = c(0, max(medicaid_percentages$percentage, na.rm = TRUE) * 1.2),
                    labels = function(x) paste0(x, "%")) +
  labs(title = "Medicaid Enrollment Among Different Groups of Adult Children Caregivers",
       subtitle = "Percentage enrolled in Medicaid",
       x = "",
       y = "Percentage") +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.title.y = element_text(size = 12),
    panel.grid.major.x = element_blank(),
    aspect.ratio = 0.6
  )
```

## Comparing Caregiving Hours Across Groups

```{r caregiving-hours, fig.width=10, fig.height=6}
# Extract caregiving hours data for visualization
extract_caregiving_hours <- function(table_data) {
  group_name <- attr(table_data, "group_name")
  
  hours_data <- table_data %>%
    filter(characteristic == "Caregiving hours") %>%
    mutate(group = group_name,
           hours = count) %>%
    select(group, level, hours)
  
  return(hours_data)
}

# Combine caregiving hours data from all tables
caregiving_hours_data <- bind_rows(
  extract_caregiving_hours(table1),
  extract_caregiving_hours(table2),
  extract_caregiving_hours(table3)
)

# Convert to long format for plotting
caregiving_hours_long <- caregiving_hours_data %>%
  pivot_wider(names_from = level, values_from = hours) %>%
  rename(ADL = `Mean ADL hours/day`, IADL = `Mean IADL hours/day`) %>%
  pivot_longer(cols = c(ADL, IADL), names_to = "hours_type", values_to = "hours")

# Plot caregiving hours comparison
ggplot(caregiving_hours_long, aes(x = group, y = hours, fill = hours_type)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  geom_text(aes(label = round(hours, 1)), 
            position = position_dodge(width = 0.7), 
            vjust = -0.5, 
            size = 4) +
  scale_y_continuous(limits = c(0, max(caregiving_hours_long$hours, na.rm = TRUE) * 1.2)) +
  labs(title = "Caregiving Hours Among Different Groups of Adult Children Caregivers",
       subtitle = "Mean hours per day spent on ADL and IADL assistance",
       x = "",
       y = "Hours per day",
       fill = "Care Type") +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.title.y = element_text(size = 12),
    panel.grid.major.x = element_blank(),
    aspect.ratio = 0.6  # Adjust this value to make the plot taller
  )
```

## Discussion

This analysis reveals several important patterns among adult children caregivers in the HEPESE dataset:

1. **Financial strain:** Adult children who moved in specifically for caregiving purposes appear to experience higher rates of financial strain compared to general co-residing caregivers. This suggests that the motivation behind co-residence may influence financial outcomes.

2. **Caregiving intensity:** The data shows variation in caregiving intensity (measured in hours spent on ADL and IADL assistance) across different groups, with those who moved in specifically for caregiving spending more hours on caregiving tasks.

3. **Demographic differences:** There are notable demographic differences between adult children who moved in specifically for caregiving versus those who co-reside for other reasons, particularly in terms of gender distribution, marital status, and income levels.

4. **Household headship:** The pattern of household headship varies across groups, which may have implications for financial decision-making, resource allocation, and caregiver autonomy within the household.

These findings have important implications for policy and practice, particularly in the development of supportive programs for Mexican-American family caregivers. Understanding the unique characteristics and needs of different types of adult child caregivers can help in designing more targeted interventions to support their caregiving roles while minimizing financial strain.

## Limitations

Several limitations should be noted when interpreting these results:

1. The data is cross-sectional, limiting our ability to establish causal relationships between co-residence patterns and financial outcomes.

2. The sample sizes for some subgroups are relatively small, which may affect the reliability of percentage estimates.

3. The HEPESE data is specific to Mexican-Americans in southwestern states and may not generalize to other Latino populations or regions.

4. Self-reported measures of financial strain and caregiving hours may be subject to recall bias or social desirability effects.

## Conclusion

This analysis provides important insights into the characteristics of Mexican-American adult children who provide care to their aging parents. The findings highlight the heterogeneity within this population and the importance of considering different pathways to co-residence when developing policies and programs to support family caregivers.

Future research should explore these patterns longitudinally to better understand how co-residence arrangements evolve over time and their long-term implications for caregiver financial well-being.